Preparing for Potential Closure of European Airspaces due to Re-entering Space Objects

### 2025-05-01 paper.qmd
Pass through abstract + I Introduction

---
title: "Preparing for Potential Closure of European Airspaces due to Reentering Space Objects"

@@

author:
  - id: es
    name: Enrico Spinielli
    affiliations:
      - ref: aiu
    orcid: 0000-0001-8584-9131
    email: enrico.spinielli@eurocontrol.int
    attributes:
      corresponding: true
  - name: Rainer Koelle
    affiliations: 
      - ref: aiu
  - name: Quinten Goens
    affiliations: 
      - ref: aiu
affiliations:
  - id: aiu
    name: EUROCONTROL
    department: Aviation Intelligence Unit
    city: Brussels
    country: Belgium

@@

# Abstract
The risk of collision between uncontrolled re-entering space debris and aircraft
is  increasing due to both ever-growing number of airborne flights and 
of orbiting bodies (and relevant re-entries). 
National and international authorities need to balance safety and economic
impact of adopted measures should a reentry event be announced, usually on short
notice.  
We employ hourly and daily spatial density of flight calculated from
trajectories as recorded by EUROCONTROL's Network Manager and
binned over Uber's H3 gridding system.  
Impact risk probabilities are then calculated based on previous theoretical
work but using these flights trajectories instead of proxies such as
population density or ADS-B/Mode S message counts.
These densities combined with a buffered area along the eventual path of
re-entry, can provide risk impact probabilities to authorities as an aid
for deciding on safety measures such as airspace closures and/or re-routings.

@@

The ever-growing forecast number of flights, see @fig-forecast, and the modern
race to commercialisation of space call for a reevaluation of the risk of
collision and casuality due to uncontrolled debris reentries.
Oversighting authorities need methodologies and tools to assess which 
safety measures are the most adequate to minimize casualties and economic
impact.

Surging number of flights in Europe [@forecastUpdate2025] and in the
world, but more importantly under development/deployment megaconstellations
of satellites on \acr{LEO} such as already active Starlink and OneWeb or
planned ones like Kuiper, Starshield, Xingwang et al. will add thousand
of orbiting objects with a short operational life some of which will 
inevitably reenter Earth atmosphere in an uncontrolled manner, cfr. @tbl-megaconstellations [@megaconstellations].

This paper develops a practical and straightforward pipeline to rapidly
assess the risk from uncontrolled reentry of space debris for the European
region.
Apart from the general risk assessment of potential reentries, we analyze
a soon to happen reentry of debris from a 1972 failed USSR mission to Venus.

@@

Debris that reenter the Earth’s atmosphere via drag-induced decay will fly
circular orbits prior to reentry even if initially on elliptical
ones.[^circular-orbit]

The angular position, $\theta$, of the reentry point from the ascending node, 
see @sec-orbitals, is uniformly distributed for uncontrolled rentries,
$\sigma(\theta) = 1 / (2\pi)$. This is due to uncertainties in atmospheric drag
and the gravitational effects of the Equatorial bulge.
Hence we assume that the impact longitude, $\lambda$, is uniformely distributed.

@@

If we consider historical debris reentry data, we can numerically derive
the debris impact probability as a weighting function by summing up and averaging
all single densities:

@@

From [@gcat] we extracted 1968 debris reentries during 25 calendar years from 
1st Jan 2000 till 31st Dec 2024.
The result of combining all the U-shaped densities can be seen in @fig-weighting.

@@

Weighting function derived from 25 years of reentries as an
empirical approximation of $\sigma(\varphi)$. 
:::

@@

## Use Case: KOSMOS 482 Descent Craft reentry

In few days (at the time of writing this paper) at 6h May 10, 2025
($\pm 2.8$ days), the KOSMOS 482 Descent Craft should reenter the
Earth's atmosphere [@kosmos482reentry].
This is supposedly the lander module from a 1972 failed Soviet Venera mission
[@SpaceReviewKosmos] of \qty{480}{kg} weight and circa \qty{1}{m} diameter.
Being supposed to survive Venus's atmosphere which is much denser and hotter than Earth's atmosphere (surface pressure 93 times that of Earth and temperatures reaching \qty{467}{degreesCelsius} (\qty{872}{degreesFarenheit})).
So it could re-enter Earth intact and not break into smaller fragments some of which could burn up.

@@

Once the reentry track is known more or less accurately, the relevant
H3 cells can be

@@

Casuality for KOSMOS 482 Descent Craft reentry. 
:::

@@

We show how National and/or International authorities could prepare for the more and more likely and frequent occurrence of uncontrolled reentry of space objects that could pose a safety concern on civilian aviation.
The techniques we implemented on realistic data and sound algorithms can contribute to the bag of tools in the hands of technical and political entities aiming at guaranteeing the utmost safety and minimal economic disruption to civil aviation in Europe.

@@

The angular position, $\theta$, of the reentry point from the ascending node
in the orbit plane is $\theta = \omega + \nu$.
It is related to the latitude, $\varphi$, via the Cartesian coordinate $k$,
$k = R \sin(\varphi) = R \sin(\theta)\sin(i)$, or

>>>>

---
title: "Preparing for Potential Closure of European Airspaces due to Re-entering Space Objects"

@@

author:
  - id: es
    name: Enrico Spinielli
    affiliations:
      - ref: aiu
    orcid: 0000-0001-8584-9131
    email: enrico.spinielli@eurocontrol.int
    attributes:
      corresponding: true
  - name: Rainer Koelle
    affiliations: 
      - ref: aiu
    orcid: 0000-0003-2810-1404
    email: rainer.koelle@eurocontrol.int
  - name: Quinten Goens
    affiliations: 
      - ref: aiu
    orcid: 0000-0003-2187-1924
    email: quinten.goens@eurocontrol.int
affiliations:
  - id: aiu
    name: EUROCONTROL
    department: Aviation Intelligence Unit
    city: Brussels
    country: Belgium

@@

# Abstract
The risk of collision between uncontrolled re-entering space debris and aircraft
is  increasing due to both ever-growing number of airborne flights and 
of orbiting bodies (and relevant re-entries). 
National and international authorities need to balance safety and economic
impact of adopted measures should a re-entry event be announced, usually on short
notice.  
We employ hourly and daily spatial flight densities calculated from
trajectories as recorded by EUROCONTROL's Network Manager and
binned with the Uber H3 geospatial indexing system.  
Impact risk probabilities are then calculated based on previous theoretical
work. This work enhances the model by employing using actual flown flight trajectories as a basis for time-varying flight densities rather than relying on proxies such as the population density or ADS-B/Mode S message counts.
These densities combined with a buffered area along the eventual path of
re-entry, can provide mid-air impact probabilities to authorities as an aid
for deciding on safety measures such as airspace closures and/or re-routings.

@@

With forecasts predicting a steadily growing number of flights, see @fig-forecast, and the modern
race to commercialise space call for a re-evaluation of the risk of
collision and casualties due to uncontrolled debris re-entries.
Overseeing authorities need methodologies and tools to assess which 
safety measures are the most adequate to minimize casualties and economic
impact.

Surging number of flights in Europe [@forecastUpdate2025] and in the
world, but more importantly under development/deployment mega-constellations
of satellites in \acr{LEO}, such as already active Starlink and OneWeb or
planned ones like Kuiper, Starshield, Xingwang et al. will add thousand
of orbiting objects with a short operational life. Some of these will 
inevitably re-enter the Earth's atmosphere in an uncontrolled manner, cfr. @tbl-megaconstellations [@megaconstellations].

This paper develops a practical and straightforward pipeline to rapidly
assess the risk from uncontrolled re-entry of space debris for the European
region.
Apart from the general risk assessment of potential re-entries, we analyse
a soon to occur re-entry of debris from a 1972 failed USSR mission to Venus.

@@

Debris that reenter the Earth’s atmosphere via drag-induced decay will fly
circular orbits prior to re-entry even if initially on elliptical
ones.[^circular-orbit]

The angular position, $\theta$, of the re-entry point from the ascending node, 
see @sec-orbitals, is uniformly distributed for uncontrolled rentries,
$\sigma(\theta) = 1 / (2\pi)$. This is due to uncertainties in atmospheric drag
and the gravitational effects of the Equatorial bulge.
Hence we assume that the impact longitude, $\lambda$, is uniformely distributed.

@@

If we consider historical debris re-entry data, we can numerically derive
the debris impact probability as a weighting function by summing up and averaging
all single densities:

@@

From [@gcat] we extracted 1968 debris re-entries during 25 calendar years from 
1st Jan 2000 till 31st Dec 2024.
The result of combining all the U-shaped densities can be seen in @fig-weighting.

@@

Weighting function derived from 25 years of re-entries as an
empirical approximation of $\sigma(\varphi)$. 
:::

@@

## Use Case: KOSMOS 482 Descent Craft re-entry

In few days (at the time of writing this paper) at 6h May 10, 2025
($\pm 2.8$ days), the KOSMOS 482 Descent Craft should reenter the
Earth's atmosphere [@kosmos482re-entry].
This is supposedly the lander module from a 1972 failed Soviet Venera mission
[@SpaceReviewKosmos] of \qty{480}{kg} weight and circa \qty{1}{m} diameter.
Being supposed to survive Venus's atmosphere which is much denser and hotter than Earth's atmosphere (surface pressure 93 times that of Earth and temperatures reaching \qty{467}{degreesCelsius} (\qty{872}{degreesFarenheit})).
So it could re-enter Earth intact and not break into smaller fragments some of which could burn up.

@@

Once the re-entry track is known more or less accurately, the relevant
H3 cells can be

@@

Casuality for KOSMOS 482 Descent Craft re-entry. 
:::

@@

We show how National and/or International authorities could prepare for the more and more likely and frequent occurrence of uncontrolled re-entry of space objects that could pose a safety concern on civilian aviation.
The techniques we implemented on realistic data and sound algorithms can contribute to the bag of tools in the hands of technical and political entities aiming at guaranteeing the utmost safety and minimal economic disruption to civil aviation in Europe.

@@

The angular position, $\theta$, of the re-entry point from the ascending node
in the orbit plane is $\theta = \omega + \nu$.
It is related to the latitude, $\varphi$, via the Cartesian coordinate $k$,
$k = R \sin(\varphi) = R \sin(\theta)\sin(i)$, or
